
## Appendix A3. Probability Tables {-}

### Normal Distribution {-}

Recall from equation (1.1) that the probability density function is defined by
$$
\mathrm{f}(y) = \frac{1}{\sigma \sqrt{2\pi }}\exp \left( -\frac{1}{2\sigma^2 }\left( y-\mu \right)^2\right)
$$
where $\mu$ and $\sigma^2$ are parameters that describe the curve. In this case, we write $y \sim N(\mu,\sigma^2)$. Straightforward calculations show that
$$
\mathrm{E}~y = \int_{-\infty}^{\infty} y \mathrm{f}(y) \, dy = \int_{-\infty}^{\infty}  y \frac{1}{\sigma \sqrt{2\pi }}\exp \left( -\frac{1}{2\sigma^2 }\left( y-\mu \right)^2 \right)  \, dy  = \mu
$$
and
$$
\mathrm{Var}~y = \int_{-\infty}^{\infty} (y-\mu)^2 \mathrm{f}(y) \, dy = \int_{-\infty}^{\infty} (y-\mu)^2 \frac{1}{\sigma \sqrt{2\pi }}\exp \left( -\frac{1}{2\sigma^2 }\left( y-\mu \right)^2 \right) \, dy = \sigma^2 .
$$
Thus, the notation $y \sim N(\mu,\sigma^2)$ is interpreted to mean the random variable is distributed normally with mean $\mu$ and variance $\sigma^2$. If $y \sim N(0,1)$, then $y$ is said to be *standard normal*.

(ref:FigA1) **Standard normal probability density function**

```{r FigA1, fig.cap='(ref:FigA1)', echo=FALSE,  out.width = "60%"}
#  FIGURE A.1 - NORMAL PLOT

par(mar=c(3.2,1,.5,.1),cex=1.2)
x <- seq(-4, 4, 0.01)
y= dnorm(x)
plot(x,y, type="l", yaxt="n", xlab="",  ylab="",las=1, cex=1.2)
mtext("y", side=1, at=1,las=1, adj=1, padj=2, cex=1.5)

x1 <- seq(0, 3, 0.1)
X <- cbind(x1,x1+.01,x1+.02,x1+.03,x1+.04,x1+.05,x1+.06,
     x1+.07,x1+.08,x1+.09)
x1 <- seq(0, 3, 0.5)
X <- cbind(x1,x1+.1,x1+.2,x1+.3,x1+.4)
x1 <- seq(0, 3, 1)
X <- cbind(x1,x1+.1,x1+.2,x1+.3,x1+.4,x1+.5,x1+.6,x1+.7,x1+.8,x1+.9)

y1 <- pnorm(X)
#round(y1, digits=4)
#write.csv(y1, file = "NormalTable2.csv")
```

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    x = 0:3,
    `0.0` = c(0.5000, 0.8413, 0.9772, 0.9987),
    `0.1` = c(0.5398, 0.8643, 0.9821, 0.9990),
    `0.2` = c(0.5793, 0.8849, 0.9861, 0.9993),
    `0.3` = c(0.6179, 0.9032, 0.9893, 0.9995),
    `0.4` = c(0.6554, 0.9192, 0.9918, 0.9997),
    `0.5` = c(0.6915, 0.9332, 0.9938, 0.9998),
    `0.6` = c(0.7257, 0.9452, 0.9953, 0.9998),
    `0.7` = c(0.7580, 0.9554, 0.9965, 0.9999),
    `0.8` = c(0.7881, 0.9641, 0.9974, 0.9999),
    `0.9` = c(0.8159, 0.9713, 0.9981, 1.0000)
  ),
  col.names = c("x", "0.0", "0.1", "0.2", "0.3", "0.4", "0.5", "0.6", "0.7", "0.8", "0.9"),
  caption = "**Standard Normal Distribution Function**"
)
```

*Notes*: Probabilities can be found by looking at the appropriate row for the lead digit and column for the decimal. For example, $\Pr ( y \le 0.1) = 0.5398$.

### Chi-Square Distribution {-}

**Chi-Square Distribution**. Several important distributions can be linked to the normal distribution. If $y_1, \ldots, y_n$ are i.i.d. random variables such that each $y_i \sim N(0,1)$, then $\sum_{i=1}^n y_i^2$ is said to have a *chi-square distribution* with parameter $n$. More generally, a random variable $w$ with probability density function
$$
\mathrm{f}(w) = \frac{2^{-k/2}}{\Gamma(k/2)} w^{k/2-1} \exp (-w/2), ~~~~~~w > 0
$$
is said to have a chi-square with $df = k$ degrees of freedom, written $w \sim \chi_k^2$. Easy calculations show that for $w \sim \chi_k^2$, we have $\mathrm{E}~w = k$ and $\mathrm{Var}~w = 2k$. In general, the degrees of freedom parameter need not be an integer, although it is for the applications of this text.


(ref:FigA2) **Several chi-square probability density functions.** Shown are curves for $df$ = 3, $df$ = 5, and $df$ = 10. Greater degrees of freedom lead to curves that are less skewed.


```{r FigA2, fig.cap='(ref:FigA2)', echo=FALSE,  out.width = "60%"}
#  FIGURE A.2 - CHI-SQUARE PLOT
par(mar=c(3.2,1,.5,.1),cex=1.2)
x <- seq(0, 20, 0.01)
y= dchisq(x,3)
plot(x,y, type="l", yaxt="n", ylab="",  las=1, cex=1.4)
mtext("w", side=1, at=8,las=1, adj=1, padj=2, cex=1.5)
y1= dchisq(x,5)
y2= dchisq(x,10)
lines(x,y1)
lines(x,y2)
text(4,.2, "df=3",cex=1.1)
text(6,.15, "df=5",cex=1.1)
text(12,.1, "df=10",cex=1.1)

prob <- t(c(seq(.6, .9, 0.1), 0.95, 0.975, 0.99, 0.995, 0.9975, 0.999))
df = t(c(1:5,seq(10,40,5),60, 120))
y1 <- matrix(0, nrow=length(df),ncol=length(prob))
for(i in 1:length(df)){
y1[i,] <- qchisq(prob,df[i]);y1[i,]}
#round(y1,digits=3)
#write.csv(y1, file = "ChiSq.csv")


```

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    df = c(1, 2, 3, 4, 5, 10, 15, 20, 25, 30, 35, 40, 60, 120),
    `0.6` = c(0.71, 1.83, 2.95, 4.04, 5.13, 10.47, 15.73, 20.95, 26.14, 31.32, 36.47, 41.62, 62.13, 123.29),
    `0.7` = c(1.07, 2.41, 3.66, 4.88, 6.06, 11.78, 17.32, 22.77, 28.17, 33.53, 38.86, 44.16, 65.23, 127.62),
    `0.8` = c(1.64, 3.22, 4.64, 5.99, 7.29, 13.44, 19.31, 25.04, 30.68, 36.25, 41.78, 47.27, 68.97, 132.81),
    `0.9` = c(2.71, 4.61, 6.25, 7.78, 9.24, 15.99, 22.31, 28.41, 34.38, 40.26, 46.06, 51.81, 74.40, 140.23),
    `0.95` = c(3.84, 5.99, 7.81, 9.49, 11.07, 18.31, 25.00, 31.41, 37.65, 43.77, 49.80, 55.76, 79.08, 146.57),
    `0.975` = c(5.02, 7.38, 9.35, 11.14, 12.83, 20.48, 27.49, 34.17, 40.65, 46.98, 53.20, 59.34, 83.30, 152.21),
    `0.99` = c(6.63, 9.21, 11.34, 13.28, 15.09, 23.21, 30.58, 37.57, 44.31, 50.89, 57.34, 63.69, 88.38, 158.95),
    `0.995` = c(7.88, 10.60, 12.84, 14.86, 16.75, 25.19, 32.80, 40.00, 46.93, 53.67, 60.27, 66.77, 91.95, 163.65),
    `0.9975` = c(9.14, 11.98, 14.32, 16.42, 18.39, 27.11, 34.95, 42.34, 49.44, 56.33, 63.08, 69.70, 95.34, 168.08),
    `0.999` = c(10.83, 13.82, 16.27, 18.47, 20.52, 29.59, 37.70, 45.31, 52.62, 59.70, 66.62, 73.40, 99.61, 173.62)
  ),
  col.names = c("df", "0.6", "0.7", "0.8", "0.9", "0.95", "0.975", "0.99", "0.995", "0.9975", "0.999"),
  caption = "**Percentiles from Several Chi-Square Distributions**"
)
```

### *t*-Distribution {-}

Suppose that $y$ and $w$ are independent with $y \sim N(0,1)$ and $w \sim \chi_k^2$. Then, the random variable $t = y / \sqrt{w/k}$ is said to have a $t$-distribution with $df = k$ degrees of freedom. The probability density function is
$$
\mathrm{f}(t) = \frac{\Gamma \left( k+ \frac{1}{2} \right)}{\Gamma(k/2)} \left( k \pi \right)^{-1/2} \left( 1 + \frac{t^2}{k} \right)^{-(k+1/2)}, ~~~~~~-\infty < t < \infty
$$
This has mean 0, for $k > 1$, and variance $k/(k-2)$ for $k > 2$.

(ref:FigA3) **Several $t$-distribution probability density functions.** The $t$-distribution with $df = \infty$ is the standard normal distribution. Shown are curves for $df$ = 1, $df$ = 5 (not labeled), and $df$ = ∞. A lower $df$ means fatter tails.


```{r FigA3, fig.cap='(ref:FigA3)', echo=FALSE,  out.width = "60%"}
#  FIGURE A.3 - t-DISTRIBUTION PLOT

par(mar=c(3.2,1,.5,.1),cex=1.2)
x <- seq(-4, 4, 0.01)
y= dnorm(x)
plot(x,y, type="l", yaxt="n", ylab="",  las=1, cex=1.4)
#mtext("t-Distribution Density",  side=2, at=.435,las=1.5, adj=.2,cex=1.8)
mtext("t", side=1, at=1,las=1, adj=1, padj=2, cex=1.5)
y1= dt(x,1)
y2= dt(x,5)
y3= dt(x,10)
lines(x,y1)
lines(x,y2)
#lines(x,y3)
text(0,.1, "df=1",cex=1.1)
#text(6,.15, "df=5",cex=1.1)
text(2.5,.38, "df=infinity",cex=1.1)
arrows(1.2, .38, 0.33,.38, code=2, angle=10, length=0.2)
arrows(-0.6, .1, -1.4,.1, code=2, angle=10, length=0.2)


prob <- t(c(seq(.6, .9, 0.1), 0.95, 0.975, 0.99, 0.995, 0.9975, 0.999))
df = t(c(1:5,seq(10,40,5),60, 120))
y1 <- matrix(0, nrow=length(df),ncol=length(prob))
for(i in 1:length(df)){
y1[i,] <- qt(prob,df[i]);y1[i,]}
y1 <- rbind(y1,qnorm(prob))
#round(y1,digits=3)
#write.csv(y1, file = "t.csv")


```

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    df = c(1, 2, 3, 4, 5, 10, 15, 20, 25, 30, 35, 40, 60, 120, "∞"),
    `0.6` = c(0.325, 0.289, 0.277, 0.271, 0.267, 0.260, 0.258, 0.257, 0.256, 0.256, 0.255, 0.255, 0.254, 0.254, 0.253),
    `0.7` = c(0.727, 0.617, 0.584, 0.569, 0.559, 0.542, 0.536, 0.533, 0.531, 0.530, 0.529, 0.529, 0.527, 0.526, 0.524),
    `0.8` = c(1.376, 1.061, 0.978, 0.941, 0.920, 0.879, 0.866, 0.860, 0.856, 0.854, 0.852, 0.851, 0.848, 0.845, 0.842),
    `0.9` = c(3.078, 1.886, 1.638, 1.533, 1.476, 1.372, 1.341, 1.325, 1.316, 1.310, 1.306, 1.303, 1.296, 1.289, 1.282),
    `0.95` = c(6.314, 2.920, 2.353, 2.132, 2.015, 1.812, 1.753, 1.725, 1.708, 1.697, 1.690, 1.684, 1.671, 1.658, 1.645),
    `0.975` = c(12.706, 4.303, 3.182, 2.776, 2.571, 2.228, 2.131, 2.086, 2.060, 2.042, 2.030, 2.021, 2.000, 1.980, 1.960),
    `0.99` = c(31.821, 6.965, 4.541, 3.747, 3.365, 2.764, 2.602, 2.528, 2.485, 2.457, 2.438, 2.423, 2.390, 2.358, 2.326),
    `0.995` = c(63.657, 9.925, 5.841, 4.604, 4.032, 3.169, 2.947, 2.845, 2.787, 2.750, 2.724, 2.704, 2.660, 2.617, 2.576),
    `0.9975` = c(127.321, 14.089, 7.453, 5.598, 4.773, 3.581, 3.286, 3.153, 3.078, 3.030, 2.996, 2.971, 2.915, 2.860, 2.807),
    `0.999` = c(318.309, 22.327, 10.215, 7.173, 5.893, 4.144, 3.733, 3.552, 3.450, 3.385, 3.340, 3.307, 3.232, 3.160, 3.090)
  ),
  col.names = c("df", "0.6", "0.7", "0.8", "0.9", "0.95", "0.975", "0.99", "0.995", "0.9975", "0.999"),
  caption = "**Percentiles from Several $t-$Distributions**"
)
```

### *F*-Distribution {-}

Suppose that $w_1$ and $w_2$ are independent with distributions $w_1 \sim \chi_m^2$ and $w_2 \sim \chi_n^2$. Then, the random variable $F = (w_1/m) / (w_2/n)$ has an $F$-distribution with parameters $df_1 = m$ and $df_2 = n$, respectively. The probability density function is
$$
\mathrm{f}(y) = \frac{\Gamma \left(\frac{m+n}{2} \right)}{\Gamma(m/2)\Gamma(n/2)} \left( \frac{m}{n} \right)^{m/2} \frac{y^{(m-2)/2}} {\left( 1 + \frac{m}{n}y \right)^{m+n+2}} , ~~~~~~y > 0
$$
This has mean $n/(n-2)$, for $n > 2$, and variance $2n^2(m+n-2)/[m(n-2)^2(n-4)]$ for $n > 4$.



(ref:FigA4) **Several $F$-distribution probability density functions**. Shown are curves for (i) $df_1$ = 1, $df_2$ = 5, (ii) $df_1$ = 5, $df_2$ = 1 (not labeled), and (iii) $df_1$ = 60, $df_2$ = 60. As $df_2$ tends to $\infty$, the $F$-distribution tends to a chi-square distribution.

```{r FigA4, fig.cap='(ref:FigA4)', echo=FALSE,  out.width = "60%"}
#  FIGURE A.4 - F-DISTRIBUTION PLOT


par(mar=c(3.2,1,.5,.1),cex=1.2)
x <- seq(0, 4, 0.01)
y= df(x,60,60)
plot(x,y, type="l", yaxt="n", ylab="",  las=1, cex=1.4)
#mtext("F-Distribution Density", side=2, at=1.17,las=1.3, adj=.2,cex=1.8)
mtext("y", side=1, at=1.5,las=1, adj=1, padj=2, cex=1.5)
#y1= df(x,2,25)
y2= df(x,5,1)
y3= df(x,1,5)
#lines(x,y1)
lines(x,y2)
lines(x,y3)
text(1.7,.8, expression(df[1]==60),cex=1.1)
text(1.7,.65, expression(df[2]==60),cex=1.1)
text(0.35,1.5, expression(df[1]==1),cex=1.1)
text(0.35,1.35, expression(df[2]==5),cex=1.1)


df1 = t(c(1:5,seq(10,40,5),60, 120))
df2 = t(c(1,3,5,seq(10,40,10),60, 120))
y1 <- matrix(0, nrow=length(df1),ncol=length(df2))
for(i in 1:length(df1)){
y1[i,] <- qf(0.95,df1[i],df2);y1[i,]}
#round(y1,digits=3)
#write.csv(y1, file = "F.csv")

```


```{r, echo=FALSE}
knitr::kable(
  data.frame(
    df_1 = c(1, 2, 3, 4, 5, 10, 15, 20, 25, 30, 35, 40, 60, 120),
    `1` = c(161.45, 199.50, 215.71, 224.58, 230.16, 241.88, 245.95, 248.01, 249.26, 250.10, 250.69, 251.14, 252.20, 253.25),
    `3` = c(10.13, 9.55, 9.28, 9.12, 9.01, 8.79, 8.70, 8.66, 8.63, 8.62, 8.60, 8.59, 8.57, 8.55),
    `5` = c(6.61, 5.79, 5.41, 5.19, 5.05, 4.74, 4.62, 4.56, 4.52, 4.50, 4.48, 4.46, 4.43, 4.40),
    `10` = c(4.96, 4.10, 3.71, 3.48, 3.33, 2.98, 2.85, 2.77, 2.73, 2.70, 2.68, 2.66, 2.62, 2.58),
    `20` = c(4.35, 3.49, 3.10, 2.87, 2.71, 2.35, 2.20, 2.12, 2.07, 2.04, 2.01, 1.99, 1.95, 1.90),
    `30` = c(4.17, 3.32, 2.92, 2.69, 2.53, 2.16, 2.01, 1.93, 1.88, 1.84, 1.81, 1.79, 1.74, 1.68),
    `40` = c(4.08, 3.23, 2.84, 2.61, 2.45, 2.08, 1.92, 1.84, 1.78, 1.74, 1.72, 1.69, 1.64, 1.58),
    `60` = c(4.00, 3.15, 2.76, 2.53, 2.37, 1.99, 1.84, 1.75, 1.69, 1.65, 1.62, 1.59, 1.53, 1.47),
    `120` = c(3.92, 3.07, 2.68, 2.45, 2.29, 1.91, 1.75, 1.66, 1.60, 1.55, 1.52, 1.50, 1.43, 1.35)
  ),
  col.names = c("$df_1$", "1", "3", "5", "10", "20", "30", "40", "60", "120"),
  caption = "**Percentiles from Several $F-$Distributions**"
)
```

