
<!-- \setcounter{page}{0} -->

<!-- \backmatter -->

<!-- \tableofcontents -->

<!-- \renewcommand{\thechapter}{\Roman{chapter}} -->
<!-- \setcounter{chapter}{0} -->

# Illustrative Statistical Code {.numbered #CIllStatCode}

<!-- \setcounter{chapter}{11} -->

<!-- \setcounter{section}{-1}  -->

## Preliminary Code {#C0IllStatCode}

Here is a list of packages used in this book. You will want to install them prior to testing out the code in this book.


```{verbatim child = './../References/Packages.Rmd'}
```

```{r ref.label = 'TableGenerate', eval = FALSE}
```

```{r ref.label = 'GammaParetoSetup', eval = FALSE}
```


\newpage


```{r child = './../OnlineSupp/Chap1/Code/Ch1CodeSummary.Rmd', eval = Ch1PdfEval}
```



\newpage


```{r child = './../OnlineSupp/Chap2/Code/Ch2CodeSummary.Rmd', eval = Ch2PdfEval}
```



\newpage


```{r child = './../OnlineSupp/Chap3/Code/Ch3CodeSummary.Rmd', eval = Ch3PdfEval}
```


\newpage


```{r child = './../OnlineSupp/Chap4/Code/Ch4CodeSummary.Rmd', eval = Ch4PdfEval}
```


\newpage

```{r child = './../OnlineSupp/Chap5/Code/Ch5CodeSummary.Rmd', eval = Ch5PdfEval}
```


\newpage

## Chapter Seven Code {#C7IllStatCode}


### Example 7.1.1. Simulate Data {.numbered}

```{r ref.label = 'Example711DataGen', eval = FALSE}
```

### Example 7.1.1. Basic Simulation VaR Optimization {.numbered}

```{r ref.label = 'Example711Optimization', eval = FALSE}
```

### Example 7.1.1. Basic Simulation Optimization Using the Distribution Function {.numbered}

```{r ref.label = 'Example711OptDF', eval = FALSE}
```

### Example 7.1.2. Basic Simulation Optimization Using the Smoothed Distribution Function {.numbered}

```{r ref.label = 'Example712DFsmooth', eval = FALSE}
```

### Example 7.1.3. Optimize Using ES {.numbered}

```{r ref.label = 'Example713OptCTE', eval = FALSE}
```

### Example 7.1.3. Optimize Using Display (7.5) {.numbered}

```{r ref.label = 'Example713Opt75', eval = FALSE}
```

### Example 7.1.3. Optimize Using Deterministic Approach {.numbered}

```{r ref.label = 'Example713Det', eval = FALSE}
```

```{r ref.label = 'Example713DetA', eval = FALSE}
```

```{r ref.label = 'Example713DetB', eval = FALSE}
```

### Example 7.1.4. Optimal Quotas Quantile Regression {.numbered}

```{r ref.label = 'Example714QuanReg', eval = FALSE}
```

```{r ref.label = 'Example714QuanRegA', eval = FALSE}
```

### Section 7.1.5. Optimalization Comparison {.numbered}

```{r ref.label = 'Section715Compare', eval = FALSE}
```

### Example 7.2.1. Set-Up {.numbered}

```{r ref.label = 'Example721SetUp', eval = FALSE}
```

### Example 7.2.1. Minimizing ES over Scenarios {.numbered}

```{r ref.label = 'Example721MinScenario', eval = FALSE}
```

### Example 7.2.1. Minimizing ES, Averaging over Scenarios {.numbered}

```{r ref.label = 'Example721AvgScenario', eval = FALSE}
```

### Exercise 7.0.  {.numbered}

```{r ref.label = 'Exercise70', eval = FALSE}
```

```{r ref.label = 'Exercise70A', eval = FALSE}
```

### Appendix. ANU Risk Distribution {.numbered}


```{r ref.label = 'dataGenProcedure', eval = FALSE}
```

```{r ref.label = 'ANURiskDistn', eval = FALSE}
```


\newpage

## Chapter Eight Code {.numbered #C8IllStatCode}

### Section 8.1.2. Member 2 ES Optimization {.numbered}

```{r ref.label = 'Section812CTE', eval = FALSE}
```

```{r ref.label = 'Section812CTEA', eval = FALSE}
```

### Section 8.1.3. Fund ES Optimization {.numbered}

```{r ref.label = 'Section813Opt', eval = FALSE}
```

```{r ref.label = 'Section813OptA', eval = FALSE}
```

### Section 8.1.4. Reinsurer ES Optimization {.numbered}

```{r ref.label = 'Section814Opt', eval = FALSE}
```

### Section 8.2.1. Including Property Risk {.numbered}

```{r ref.label = 'Section821Property', eval = FALSE}
```


### Section 8.2.2. Varying Alpha {.numbered}

```{r ref.label = 'Section822VaryAlpha', eval = FALSE}
```

### Section 8.2.3. Range Value at Risk {.numbered}

```{r ref.label = 'Section823RVaR', eval = FALSE}
```

### Section 8.2.4. Claim Level Retention {.numbered}

```{r ref.label = 'Section824Claim', eval = FALSE}
```

### Appendix. Generate Property Fund Data {.numbered}

```{r ref.label = 'Chap8PropFundData', eval = FALSE}
```

### Appendix. Set-Up ANU Parameters {.numbered}

```{r ref.label = 'Chap8ANUSetUpParams', eval = FALSE}
```

### Appendix. ANU Data Generation {.numbered}

```{r ref.label = 'Chap8ANUdataGen', eval = FALSE}
```

\newpage

## Chapter Nine Code {.numbered #C9IllStatCode}

### Example 9.3.1. Develop Baseline Optimization {.numbered}

```{r ref.label = 'Example931Base', eval = FALSE}
```

### Example 9.3.1. Scale Two Sensitivities {.numbered}

```{r ref.label = 'Example931Sens2', eval = FALSE}
```

### Section 9.3.2. Wisconsin Base Results {.numbered}

```{r ref.label = 'Section932WiscBase', eval = FALSE}
```

### Section 9.3.2. Wisconsin Sensitivities {.numbered}

```{r ref.label = 'Section932WiscSensi', eval = FALSE}
```

### Section 9.3.3. ANU Base Results {.numbered}

```{r ref.label = 'Section933ANUBase', eval = FALSE}
```

### Section 9.3.3. ANU Sensitivities {.numbered}

```{r ref.label = 'Section933ANUSensi', eval = FALSE}
```

```{r ref.label = 'Section933ANUBoot', eval = FALSE}
```

### Section 9.4.1. Asset Allocation Sensitivities {.numbered}

```{r ref.label = './ProofTheory/Chap9/Example941RCode.Rmd', eval = FALSE}
```


\newpage

## Chapter Ten Code {.numbered #C10IllStatCode}

### Example 10.4.6. Varying the Cyber Risk Premium {.numbered}

```{r ref.label = 'Example1046', eval = FALSE}
```

\newpage

## Chapter Eleven Code {.numbered #C11IllStatCode}

### Example 11.1.1 Reinsurance Premiums {.numbered}

```{r ref.label = 'Example1111Estimation', eval = FALSE}
```

```{r ref.label = 'Example1111DepSens', eval = FALSE}
```

### Example 11.1.2 Value at Risk for an Insurance Portfolio {.numbered}

```{r ref.label = 'Example1112DepSens', eval = FALSE}
```

### Example 11.2.1. Portfolio Distribution with Gamma and Pareto Risks {.numbered}

```{r ref.label = 'Example1121', eval = FALSE}
```

### Example 11.2.2. Portfolio Distribution - Simulation Based Computations {.numbered}

```{r ref.label = 'Example1122', eval = FALSE}
```


### Example 11.3.3  {.numbered}

```{r ref.label = 'Example1133DepSens', eval = FALSE}
```

### Example 11.4.1  {.numbered}

```{r ref.label = 'Example1141DepSens', eval = FALSE}
```


# Selected Proofs and Theory Development {.numbered #CProofs}

## Chapter Two Theory Development {#C2Proofs}

### Section 2.1. Show Development of the $ES$ {.numbered}


```{r child = './../ProofTheory/Chap2/Section21ES.Rmd'}
```

### Section 2.2.3. Show Development of the $ES$ {.numbered}

```{r child = './../ProofTheory/Chap2/Section223ES.Rmd'}
```

### Section 2.2.3. Show Development of the $RVaR$ {.numbered}

```{r child = './../ProofTheory/Chap2/Section223RVaR.Rmd'}
```

<!-- ### Section 2.5. Show Development of the $ES$ Sensitivity {.numbered} -->

<!-- ```{r child = './../ProofTheory/Chap2/Section25CTESensitivity.Rmd'} -->
<!-- ``` -->

<!-- ### Example 2.5.1. Show Development of the Upper Limit Case {.numbered} -->

<!-- ```{r child = './../ProofTheory/Chap2/Example251.Rmd'} -->
<!-- ``` -->


## Chapter Three Theory Development {#C3Proofs}

### Section 3.1. Verification of Proposition 3.1 {.numbered}

```{r child = './../ProofTheory/Chap3/Proposition31.Rmd'}
```

### Section 3.1. Verification of Proposition 3.2 {.numbered}

```{r child = './../ProofTheory/Chap3/Proposition32.Rmd'}
```

### Section 3.1. Verification of Proposition 3.3 {.numbered}

```{r child = './../ProofTheory/Chap3/Proposition33.Rmd'}
```

### Section 3.1. Verification of Proposition 3.4 {.numbered}

```{r child = './../ProofTheory/Chap3/Proposition34.Rmd'}
```


## Chapter Four Theory Development {#C4Proofs}

### Section 4.1.1.1. Verification of the Optimal Retention Proportions {.numbered}

```{r child = './../ProofTheory/Chap4/Section4111OptQuotaShare.Rmd'}
```

### Section 4.1.2.2. Verification of the Optimal Excess of Loss Limits {.numbered}

```{r child = './../ProofTheory/Chap4/Section4122OptExLoss.Rmd'}
```

### Section 4.2.2. Check the Limited Cross Product Expectation Equation (4.6) {.numbered}

```{r child = './../ProofTheory/Chap4/Section422CrossProd.Rmd'}
```

### Section 4.3.2. Quota Share with Parameter Constraints {.numbered}

```{r child = './../ProofTheory/Chap4/Section432QuotaShare.Rmd'}
```

### Section 4.3.3. Verification of Equation (4.7) {.numbered}

```{r child = './../ProofTheory/Chap4/Section433Eq47.Rmd'}
```

### Section 4.3.3. Independence Case Details {.numbered}

```{r child = './../ProofTheory/Chap4/Section433IndRisks.Rmd'}
```



### Section 4.4.1. Verification of the Convexity {.numbered}

```{r child = './../ProofTheory/Chap4/Section441Convexity.Rmd'}
```

### Example 4.4.2. Verification of the Derivative of the Objective Function {.numbered}

```{r child = './../ProofTheory/Chap4/Example442.Rmd'}
```



### Example 4.5.2. Development of the Exchange for an Exponential Utility Function {.numbered}

```{r child = './../ProofTheory/Chap4/Example452.Rmd'}
```

## Chapter Five Theory Development {#C5Proofs}

### Section 5.2.1. Verification of the Excess of Loss Distribution Function {.numbered}

```{r child = './../ProofTheory/Chap5/Section521ExcessLoss.Rmd'}
```

### Section 5.2.2. Verification of the Excess of Loss Distribution Function Partial Derivative {.numbered}

```{r child = './../ProofTheory/Chap5/Section522ExLossPartial.Rmd'}
```

### Section 5.3.1. Verification of the $VaR$ Derivative Expressions {.numbered}

```{r child = './../ProofTheory/Chap5/Section531ExLossVaRDeriv.Rmd'}
```

### Section 5.3.1. Verification of the $ES$ Derivative Expressions {.numbered}

```{r child = './../ProofTheory/Chap5/Section531ExLossESDeriv.Rmd'}
```

### Section 5.3.1. Verification of the $ES$ Excess of Loss Derivative Expressions {.numbered}

```{r child = './../ProofTheory/Chap5/Section531ExLossESDeriv2.Rmd'}
```

### Section 5.4.2. Verification of the Bounds for Active Constraints{.numbered}

```{r child = './../ProofTheory/Chap5/Section542Active.Rmd'}
```

## Chapter Seven Theory Development {#C7Proofs}

### Section 7.1.1. Verification of the VaR as a Solution of Display (7.3) {.numbered}

```{r child = './../ProofTheory/Chap7/Section711Eq73.Rmd'}
```

### Section 7.1.2. Solution of the ES Minimization Problem {.numbered}

```{r child = './../ProofTheory/Chap7/Section712CTE.Rmd'}
```

### Section 7.1.3. Solution of the Quantile Minimization Problem {.numbered}

```{r child = './../ProofTheory/Chap7/Section713Quantile.Rmd'}
```

### Section 7.1.3. Confirm the Relationship between Quantile Regression and ES Objective Functions {.numbered}

```{r child = './../ProofTheory/Chap7/Section713QuantileRelate.Rmd'}
```

### Section 7.3.1. Confirm Display (7.13) {.numbered}

```{r child = './../ProofTheory/Chap7/Section731Display713.Rmd'}
```

### Section 7.3.1. Confirm Equation (7.14) {.numbered}

```{r child = './../ProofTheory/Chap7/Section731Eq714.Rmd'}
```

### Section 7.3.1. Confirm the Scaling Factor to Maintain the Budget Constraint {.numbered}

```{r child = './../ProofTheory/Chap7/Section731Scale.Rmd'}
```


## Chapter Nine Theory Development {#C9Proofs}

### Example 9.2.4. Check the VaR Sensitivity {.numbered}

```{r child = './../ProofTheory/Chap9/Example924.Rmd'}
```

### Section 9.4.1. Verification of Asset Allocation Sensitivities {.numbered}

```{r child = './../ProofTheory/Chap9/Section941.Rmd'}
```

## Chapter Ten Theory Development {#C10Proofs}

### Example 10.1.1. Verification that the Optimal d=0 {.numbered}

```{r child = './../ProofTheory/Chap10/Example1011.Rmd'}
```

### Section 10.2. Proof the Single Risk Retention Results {.numbered}

```{r child = './../ProofTheory/Chap10/Section102.Rmd'}
```

### Example 10.3.1. Confirm the $RM^2$ {.numbered}

```{r child = './../ProofTheory/Chap10/Example1031.Rmd'}
```


### Example 10.3.2. Check the $VaR$ Balance {.numbered}

```{r child = './../ProofTheory/Chap10/Example1032.Rmd'}
```

### Example 10.3.4. Confirm the $RM^2$ for Example 10.3.4 {.numbered}

```{r child = './../ProofTheory/Chap10/Example1034.Rmd'}
```

### Section 10.4.1. Confirm the $RM^2$ for Variance Risk Measure with Simulation {.numbered}

```{r child = './../ProofTheory/Chap10/Section1041.Rmd'}
```

### Section 10.4.2. Confirm the Conditions for No Risk Retention {.numbered}

```{r child = './../ProofTheory/Chap10/Section1042.Rmd'}
```


## Chapter Eleven Theory Development {#C11Proofs}


### Example 11.2.1. Development of Two Risk Portfolio {.numbered}

```{r child = './../ProofTheory/Chap11/Section1121.Rmd'}
```


### Example 11.3.3. Development of the RVaR Sensitivity {.numbered}

```{r child = './../ProofTheory/Chap11/Section1133.Rmd'}
```


# Selected Exercise Solutions {.numbered #CSolutions}

## Chapter One Exercise Solutions {#C1Solutions}

### Exercise 1.1. Solution {.numbered}

```{r ref.label = 'Exercise11', eval = FALSE}
```

```{r ref.label = 'Exercise11FigAa', eval = FALSE}
```

```{r ref.label = 'Exercise11Soln', eval = FALSE}
```

```{r child = './../ProofTheory/Chap1/Exercise11Soln.Rmd'}
```

### Exercise 1.2. Solution {.numbered}

```{r ref.label = 'Exercise12Fig', eval = FALSE}
```

```{r child = './../ProofTheory/Chap1/Exercise12Soln.Rmd'}
```


## Chapter Two Exercise Solutions {#C2Solutions}

### Exercise 2.1.1. Value at Risk ($VaR$) for Pareto and Gamma Distributions {.numbered}

```{r ref.label = 'Exer211', eval = FALSE}
```

```{r ref.label = 'Exer211B', eval = FALSE}
```

### Exercise 2.1.2. Comparing Parametric $RVaR$ Measures {.numbered}

```{r ref.label = 'Exer212', eval = FALSE}
```

```{r ref.label = 'WebGL', eval = FALSE}
```


### Exercise 2.1.3. Risk Measures for the Sum of Two Risks {.numbered}

```{r ref.label = 'Exer213', eval = FALSE}
```

### Exercise 2.2.1. Risk Retention for the Sum of Two Risks {.numbered}

```{r ref.label = 'Exer221', eval = FALSE}
```

### Exercise 2.2.2. Portfolio Management and Simulation {.numbered}

```{r ref.label = 'Exer222SetUp', eval = FALSE}
```

```{r ref.label = 'Exer222', eval = FALSE}
```

### Exercise 2.3.1. Risk Retention Changes for the Sum of Two Risks {.numbered}

```{r ref.label = 'Exer231', eval = FALSE}
```


## Chapter Three Exercise Solutions {#C3Solutions}

### Exercise 3.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap3/Exercise31Soln.Rmd'}
```

### Exercise 3.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap3/Exercise32Soln.Rmd'}
```


### Exercise 3.3. Alternative Training and Testing Periods for Portfolio of Insurance Stock Returns {.numbered}

```{r ref.label = 'Exer33', eval = FALSE}
```

```{r ref.label = 'Exercise33', eval = FALSE}
```

```{r child = './../ProofTheory/Chap3/Exercise33Soln.Rmd'}
```

## Chapter Four Exercise Solutions {#C4Solutions}

### Exercise 4.3.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise431Soln.Rmd'}
```

### Exercise 4.4.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise441Soln.Rmd'}
```

### Exercise 4.5.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise451Soln.Rmd'}
```

### Exercise 4.5.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise452Soln.Rmd'}
```

### Exercise 4.5.3 Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise453Soln.Rmd'}
```


### Exercise 4.5.4 Solution {.numbered}

```{r ref.label = 'Exercise454', eval = FALSE}
```


```{r ref.label = 'Exercise454', eval = FALSE}
```

```{r ref.label = 'Exercise4541', eval = FALSE}
```

```{r ref.label = 'Exercise455', eval = FALSE}
```


### Exercise 4.5.4 Solution {.numbered}

### Exercise 4.5.6. Solution {.numbered}

```{r child = './../ProofTheory/Chap4/Exercise456Soln.Rmd'}
```

## Chapter Five Exercise Solutions {#C5Solutions}

### Exercise 5.2.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise521Soln.Rmd'}
```

### Exercise 5.2.3. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise523Soln.Rmd'}
```


### Exercise 5.3.1. Risk Sensitivity the Sum of Two Risks {.numbered}

```{r ref.label = 'Exer531', eval = FALSE}
```


### Exercise 5.3.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise532Soln.Rmd'}
```


### Exercise 5.3.3. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise533Soln.Rmd'}
```


### Exercise 5.3.4. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise534Soln.Rmd'}
```

### Exercise 5.4.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise541Soln.Rmd'}
```

### Exercise 5.4.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise542Soln.Rmd'}
```

### Exercise 5.4.3. Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise543Soln.Rmd'}
```

### Exercise 5.4.4 Solution {.numbered}

```{r child = './../ProofTheory/Chap5/Exercise544Soln.Rmd'}
```

## Chapter Seven Exercise Solutions {#C7Solutions}

### Exercise 7.1 {.numbered}

```{r ref.label = 'Exer71', eval = FALSE}
```

## Chapter Eight Exercise Solutions {#C8Solutions}


### Exercise 8.4. Code for Solution  {.numbered}

```{r child = './../ProofTheory/Chap8/Exercise84.Rmd'}
```

```{r ref.label = 'Exercise84Soln', eval = FALSE}
```


### Exercise 8.5. Code for Solution  {.numbered}

```{r ref.label = 'Exercise85Soln', eval = FALSE}
```


## Chapter Nine Exercise Solutions {#C9Solutions}

### Exercise 9.1.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise911Soln.Rmd'}
```


### Exercise 9.1.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise912Soln.Rmd'}
```


### Exercise 9.1.3. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise913Soln.Rmd'}
```

### Exercise 9.1.4. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise914Soln.Rmd'}
```

### Exercise 9.1.5. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise915Soln.Rmd'}
```

### Exercise 9.2.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise921Soln.Rmd'}
```

### Exercise 9.2.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise922Soln.Rmd'}
```

### Exercise 9.2.3. Solution {.numbered}


```{r ref.label = 'SenAnalGVaRTab1', eval = FALSE}
```

```{r ref.label = 'Exercise923', eval = FALSE}
```

### Exercise 9.2.4. Solution {.numbered}

```{r ref.label = 'Exercise924', eval = FALSE}
```

### Exercise 9.3.1. Part (a) Solution {.numbered}

```{r ref.label = 'Exercise931A', eval = FALSE}
```

### Exercise 9.3.1. Part (b) Solution {.numbered}

```{r ref.label = 'Exercise931B', eval = FALSE}
```

### Exercise 9.3.1. Part (c) Solution {.numbered}

```{r ref.label = 'Exercise931C', eval = FALSE}
```

### Exercise 9.3.1. Part (d) Solution {.numbered}

```{r ref.label = 'Exercise931D', eval = FALSE}
```

### Exercise 9.4.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap9/Exercise941Soln.Rmd'}
```

## Chapter Ten Exercise Solutions {#C10Solutions}

### Exercise 10.1. Solution {.numbered}

```{r child = './../ProofTheory/Chap10/Exercise101Soln.Rmd'}
```


### Exercise 10.2. Solution {.numbered}

```{r child = './../ProofTheory/Chap10/Exercise102Soln.Rmd'}
```


### Exercise 10.3. Solution {.numbered}

```{r child = './../ProofTheory/Chap10/Exercise103Soln.Rmd'}
```


### Exercise 10.4. Solution {.numbered}

```{r child = './../ProofTheory/Chap10/Exercise104Soln.Rmd'}
```

